Forecasting the 2024 U.S. Presidential Election: Kamala Harris' Projected Victory

### 2024-11-04 paper/paper.qmd
commmitting newly rendered paper.qmd

# Introduction {#sec-intro}
The 2024 United States Presidential Election is anticipated to be a pivotal moment in American political history, particularly as battleground states emerge as key determinants of the electoral outcome. With candidates Donald Trump and Kamala Harris intensifying their campaigns, a thorough understanding of voter preferences in these critical regions becomes essential for forecasting election results. This paper leverages state-level polling data from Project 538 [@FiveThirtyEight] and employs the statistical programming language R [@citeR] to construct a predictive model that estimates the likelihood of victory for each candidate across several crucial battleground states, including Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin.

The primary estimand of this study is the probability of winning for either Donald Trump or Kamala Harris in these battleground states, derived from aggregated polling averages. Our analysis utilizes a Bayesian hierarchical approach combined with natural splines to accurately capture the temporal dynamics of voter sentiment, enabling us to estimate winning probabilities in these decisive areas. By comparing the predicted winning probabilities for both candidates, we can ascertain the likely winner based on prevailing voter support trends. Our findings indicate a competitive landscape, with Kamala Harris projected to secure victories in six of the seven battleground states analyzed, including key states like Michigan and Nevada, where voter sentiment appears to favor her candidacy. However, Donald Trump is also predicted to maintain a competitive edge in Arizona, showcasing the variability in voter preferences across these critical regions. The predicted winner is determined by comparing the estimated probabilities for each candidate, highlighting Harris’s overall advantage leading into the election.

Understanding these predictions is crucial as they not only forecast the likely outcome of the U.S. election but also help society and the economy to better understand the current state of the US political climate. The implications of these findings extend beyond mere predictions, providing a framework for analyzing how shifts in voter sentiment may occur as the election draws near. The remainder of this paper is organized as follows: Section @sec-data details the data and variables utilized in the analysis; Section @sec-model outlines the model setup and estimation strategies employed; Section @sec-results presents the results, including visualizations of the predicted election outcomes; and Section @sec-discussion discusses the implications and limitations of our findings.

# Data {#sec-data}
## Overview
The data for this paper contains the polling information of the 2024 Presidential Election cycle and was obtained from Project 538's polling project [@FiveThirtyEight] on October 19th, 2024. The data is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as state, start_date, end_date, transparency_score, candidate_name, and percentage of support (pct) for the appropriate candidates are included. The dataset compiles data from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the variables mentioned (see @sec-variables), we collected observations for both of the candidates. We collect observations from the dataset where the ‘candidate_name’ is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of ‘end_date’, and ‘state’, as well as the electorate support for each candidate is denoted as ‘pct’. Furthermore, since there are many pollsters present in the dataset, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high ‘numeric_grade’ and ‘transparency_score’, details of data cleaning are further explained in [Appendix -@sec-data-details].

@@

Once collected, these individual responses are combined into percentages of support. This aggregation abstracts further, treating all “Trump” responses equally, regardless of how confident or reluctant the voter might have been. In effect, the complexity of each opinion is smoothed over into a uniform data point, with many variations of opinions lost in the process.

Finally, these support percentages are fed into a model to make a predictive statement about the election outcome. The model’s goal is to determine who is likely to win, based on the assumption that the aggregated opinions in our data reflect real-world voter behavior on Election Day. However, this step relies heavily on simplification, assuming that the snapshot of opinions captured in the polls will hold steady and that all voters will follow through on their expressed preferences.

The raw polling data gives us percentages of support for each candidate in each state. To measure our outcome variable (the likelihood of winning), we convert these percentages into a binary outcome in our model (predicting Trump or Harris as the winner based on who has higher support in each state).  Our model uses these support percentages to generate a probability of each candidate winning. This transformation helps translate fluctuating polling data into a clear prediction of the likely winner, but it also involves a level of simplification, as it assumes these support percentages accurately reflect voting intentions.

@@

It is important to note that since ‘pct’ stands for percentage, there may be a misconception that the values for Trump and Harris will always add up to 100%. For instance, in the summary statistics presented in Table 1, the average support for Donald Trump is 47.801%, while the average support for Kamala Harris is 47.999%. This results in a combined average support of 95.800%, indicating that there is a significant portion of respondents who are either undecided or supporting other candidates. This highlights the competitive nature of the election and emphasizes the necessity of examining not only the percentages attributed to the main candidates but also the context of voter preferences that could influence the final outcome.

### Predictive Variables
In this section, we outline the two key predictor variables used in our analysis: 'state', which captures the geographical context of the polling data, and 'end_date', which reflects the timing of the polls relative to the election, allowing us to assess trends in voter support over time.

#### State
The state ('state') variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground statesn. Notably, Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations. "

@@

#### End Date {#sec-end-date}
End date ('end_date') represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states, highlighting peaks in polling activity around late September."

@@

### Other Variables
This section details additional variables that provide essential context for our analysis, including numeric grade, which evaluates the quality of the polls; transparency score, indicating the openness of poll methodologies; pollster, identifying the organizations conducting the polls; and start date, which denotes the initiation of the polling period for each survey.

#### Numeric Grade
Numeric Grade ('numeric_grade') represents the quality of each poll, as rated by [@FiveThirtyEight], based on factors such as methodology, transparency, and historical accuracy. In the context of our paper, numeric grade allows us to weigh the reliability of different polls. By including numeric grade as a variable in our model, we aim to account for variations in poll quality, ensuring that polls with higher ratings (indicating higher reliability) contribute more significantly to our predictions of voter support. @fig-numgrade-distribution displays the distribution of numeric grades for the polls included in our analysis, ranging from 2.5 to 3.0. Polls with higher grades are generally seen as more methodologically rigorous and transparent, which helps improve the reliability of our model’s predictions. The distribution shows several peaks, indicating that certain grade levels (around 2.7, 2.8, and 3.0) are more common among the polling data we collected.

@@

```
This variability in numeric grade highlights the diversity in polling quality within our dataset. The presence of peaks suggests that some pollsters consistently receive higher ratings, likely due to their use of more rigorous methods or transparent reporting standards. Conversely, the valleys in the distribution may indicate fewer polls from sources with lower numeric grades, reflecting a potential tendency for lower-quality polls to be less prevalent or less frequently conducted. By accounting for numeric grade, we can improve the accuracy of predictions by emphasizing data from more reliable sources. This consideration is crucial, as polling quality can impact the stability of the model’s forecasts, especially in closely contested states where polling accuracy is essential for reliable predictions.
\newpage

#### Transparency Score
Transparency Score ('transparency_score') represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data. Indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

#### Pollster
Pollster ('pollster') represents the various organizations responsible for conducting polls across key battleground states in the lead-up to the 2024 United States Presidential Election. In the context of our paper, the diversity and frequency of polling by different pollsters are essential for understanding the reliability and variance in the data. Each pollster may use different methodologies, sample populations, and data collection techniques, which can affect the results and introduce variability in polling outcomes.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar chart displaying the distribution of polls conducted by various pollsters in key battleground states. Highlights the frequency of polling activities across different organizations, with Marist and Emerson standing out as the most prolific pollsters."

@@

#### Start Date
Start date ('start_date') represents the date when each poll began collecting responses. In the context of our paper, start date is a key variable as it allows us to focus on polling data collected after July 21, 2024—the date Kamala Harris officially announced her presidential campaign. By filtering the dataset to include only polls with a start date after this announcement, we ensure that each poll reflects voter sentiment with Harris as an active candidate, enabling a more accurate comparison between her and Donald Trump. @fig-start-date-distribution displays the distribution of polling start dates from late July to mid-October 2024, showing when polling efforts began across key battleground states. The histogram reveals peaks in polling activity, particularly in mid-August, early September, and late September, indicating periods of heightened interest or campaign events that likely prompted increased polling.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states. Displays the frequency of polling activities initiated over time, with a notable peak on September 14, indicating a surge in polling efforts as the election approached."

@@

# Model {#sec-model}
The Bayesian Beta regression models constructed in this paper aim to predict the probability of victory for Donald Trump and Kamala Harris in the 2024 U.S. Presidential Election by analyzing polling data of the 7 battleground states. One Bayesian Beta regression model is constructed for each candidate, and each of them estimates the proportion of support for each candidate based on polling data, accounting for both polling trends over time, and state-level differences in voter sentiment. These models allow us to evaluate how support for each candidate fluctuates over time in key states, and predict potential electoral outcomes.

@@

## Model Summary
### Trump Model Summary
@tbl-trump-model-summary fixed effects estimates for several key terms that capture the influence of time on his support in the battleground states. The intercept, with an estimate of -0.0923, represents the baseline log-odds of Trump’s support when other predictors are held constant. This negative intercept suggests a slightly lower baseline level of support. The term nsenddatenumdfEQ42, with an estimate of 0.0803 and a statistically significant positive confidence interval (0.0179 to 0.1426), indicates an increase in support around this time period. Similarly, nsenddatenumdfEQ44 has an estimate of 0.0789 and a confidence interval above zero (0.0014 to 0.1589), also showing a positive effect on support. Other terms such as nsenddatenumdfEQ41 and nsenddatenumdfEQ43 have less significant impacts, suggesting that certain time periods were more influential than others in shaping Trump’s support. The standard errors for the fixed effects terms range from 0.0306 to 0.0676 for nsenddatenumdfEQ43. The relatively low standard error of 0.0306 on the intercept suggests that the baseline estimate is quite stable, meaning we can be more confident about this baseline support level. However, the standard error for nsenddatenumdfEQ43 (0.0676) is higher than for other terms, indicating that this estimate is less precise.
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-model-summary
#| tbl-cap: "Fixed Effects Summary for Trump Model."
#| warning: false

@@

### Harris Model Summary
@tbl-harris-model-summary similarly shows the impact of time on her support in the battleground states. The intercept for the Harris model is estimated at -0.1508, suggesting a slightly lower baseline support level compared to Trump(-0.0923). The term nsenddatenumdfEQ41 has an estimate of 0.1225 and a confidence interval 0.0553 to 0.1903, indicating a significant increase in support during this period. Another positive and significant effect is seen in nsenddatenumdfEQ43, with an estimate of 0.1191 and confidence interval from -0.0071 to 0.2417, which suggests a trend toward increased support. While other terms stay positive, they have less impacts on support. Additionally, the standard errors for fixed effects are generally smaller, ranging from 0.0279 for the intercept to 0.0631 for nsenddatenumdfEQ43. So the larger standard error for nsenddatenumdfEQ43 (0.0631) suggests that there would be some uncertainty in how this time point influences support, though the positive estimate still points toward an upward trend. In comparison to the Trump model, the standard errors in the Harris model are slightly smaller for most terms, indicating potentially more consistent support trends over time in the data analyzed.
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-model-summary
#| tbl-cap: "Fixed Effects Summary for Harris Model."
#| warning: false

# Summarize fixed effects for Harris model
harris_summary <- tidy(harris_model, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Estimate = estimate,
    `Std. Error` = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  mutate(term = gsub("[^[:alnum:] ]", "", term))  # Clean up term names

@@

## State-Level Polling Averages for Trump and Harris
@tbl-outcomevar displays overview of the percentage of support for Trump and Harris by each battleground state. It provides us with a high level understanding of the current voter dynamic. The pct results are aggregated averages of all pct data points for Trump and Harris in each state. By comparing the percentage of support for T.rump and Harris, we can determine the candidate that has a leading advantage in winning each state. For example,Trump has 49.2% support in Arizona while Harris has 46.8% support, which means Trump has won the state of Arizona.
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-stateaverage
#| tbl-cap: "Table displaying the percentage of support for Donald Trump and Kamala Harris across key battleground states in the lead-up to the 2024 United States Presidential Election. The table lists states with their respective support percentages, highlighting the competitive landscape as voters prepare to make their decisions on Election Day. Notably, Trump maintains a slight lead in states such as Arizona and Georgia, while Harris shows stronger support in states like Michigan and Nevada. This data provides insight into the current polling dynamics and potential electoral outcomes in these critical states."

@@

## Polling Averages for Trump and Harris Over Time in Battleground States
@fig-state-polling-average provides a visual comparison of the current polling support for Donald Trump and Kamala Harris across key battleground states, tracked over time from August 2024 to October 2024. Each line plot represents the average polling percentages for both candidates in a specific state, with red lines denoting Trump’s support and blue lines representing Harris’s support. Across several states, there are noticeably increased fluctuations in polling averages for both candidates closer to the election particularly from late September to October. This pattern may indicate heightened voter interest and engagement as the election date approaches which is November 4th, 2024, it could also reflect changes in public opinion due to campaign events, political debates, or other external factors.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-polling-average
#| fig-cap: "Polling Averages for Trump and Harris Over Time in Battleground States."
#| warning: false

@@

# Plot polling averages over time for each battleground state
ggplot(polling_average, aes(x = end_date)) +
  geom_line(aes(y = Average_Trump, color = "Trump")) +
  geom_line(aes(y = Average_Harris, color = "Harris")) +
  labs(title = "Polling Averages for Trump and Harris Over Time in Battleground States", 
       y = "Average Support (%)", 
       x = "Date") +
  facet_wrap(~ state) +
  scale_color_manual(values = c("Trump" = "red", "Harris" = "blue")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +  # Set date format and break frequency
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1),  # Rotate x-axis labels
    strip.text = element_text(size = 10),  # Control facet label size if needed
    panel.spacing = unit(0.5, "lines"),    # Add spacing between facets
    panel.grid.major = element_line(linewidth = 0.3)  # Adjust gridline thickness
  )

@@

Since it is difficult to conclude a definitive predictive outcome by observing the current data, we utilize our models, described in @sec-model to predict the possible outcomes by incorporating the given data. By using the current data, the model can identify the state-specific trends thus generating more accurate predictions of the outcome. Relevant prediction results are provided below.

## State-Level Probability of Trump Winning on Election Day
By incorporating our Bayesian models, we generate the predicted probabilities of winner for both Donald Trump and Kamala Harris in key battleground states, on Election Day (November 5th, 2024). @fig-state-probabilities provides a visual summary of these predicted probabilities.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-probabilities
#| fig-cap: "Probability of Trump Winning by State on Election Day."
#| warning: false

>>>>

# Introduction {#sec-intro}
The 2024 United States Presidential Election is set to be a crucial moment in American political history, with battleground states serving as pivotal determinants of the electoral outcome. As candidates Donald Trump and Kamala Harris intensify their campaigns, understanding voter preferences in these key regions is essential for accurate election forecasting. This paper utilizes state-level polling data from FiveThirtyEight [@FiveThirtyEight] and the statistical programming language R [@citeR] to develop a predictive model estimating each candidate's likelihood of victory in critical battleground states, including Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin.

The primary focus of this study is the probability of victory for Trump or Harris in these battleground states, derived from aggregated polling averages. Employing a Bayesian hierarchical model combined with natural splines, our analysis effectively captures the temporal dynamics of voter sentiment, allowing us to estimate winning probabilities in these decisive areas. Our findings reveal a competitive landscape, with Harris projected to win in six of the seven battleground states analyzed, notably in Michigan and Nevada, where support leans in her favor. However, Trump remains competitive in Arizona, illustrating the variability of voter preferences across these regions.

These predictions are critical not only for forecasting the likely outcome of the election but also for enhancing societal and economic understanding of the current political climate in the U.S. The implications of our findings extend beyond mere predictions, offering a framework for analyzing potential shifts in voter sentiment as Election Day approaches. The remainder of this paper is organized as follows: Section @sec-data details the data and variables utilized in the analysis; Section @sec-model outlines the model setup and estimation strategies employed; Section @sec-results presents the results, including visualizations of the predicted election outcomes; and Section @sec-discussion discusses the implications and limitations of our findings.

# Data {#sec-data}
## Overview
The data for this paper contains the polling information of the 2024 Presidential Election cycle and was obtained from Project 538's polling project [@FiveThirtyEight] on October 19th, 2024. The data is periodically updated throughout the presidential election campaign to reflect current and most up-to-date polling results. It consists of polling information from various pollsters such as the New York Times, ActiVote, Morning Consult, etc. For every poll conducted, relevant variables such as state, start_date, end_date, transparency_score, candidate_name, and percentage of support (**pct**) for the appropriate candidates are included. The data compiles information from all major pollsters in the US and provides us with an understanding of the current state of the presidential election.

The present analysis focuses on Donald Trump and Kamala Harris, the two leading candidates in the current presidential election race. For all the variables mentioned (see @sec-variables), we collected observations for both of the candidates. We collect observations from the data where the ‘candidate_name’ is Trump and Harris. We are interested in the support for both candidates over time, for each state, therefore, we collect observations of ‘end_date’, and ‘state’, as well as the electorate support for each candidate is denoted as ‘pct’. Furthermore, since there are many pollsters present in the data, we extracted pollsters with high-quality data that is reliable, which is denoted as having a high ‘numeric_grade’ and ‘transparency_score’, details of data cleaning are further explained in [Appendix -@sec-data-details].

@@

Once collected, these individual responses are combined into percentages of support. This aggregation abstracts further, treating all “Trump” responses equally, regardless of how confident or reluctant the voter might have been. In effect, the complexity of each opinion is smoothed over into a uniform data point, with many variations of opinions lost in the process. Finally, these support percentages are fed into a model to make a predictive statement about the election outcome. The model’s goal is to determine who is likely to win, based on the assumption that the aggregated opinions in our data reflect real-world voter behavior on Election Day. However, this step relies heavily on simplification, assuming that the snapshot of opinions captured in the polls will hold steady and that all voters will follow through on their expressed preferences. The raw polling data gives us percentages of support for each candidate in each state. To measure our outcome variable (the likelihood of winning), we convert these percentages into a binary outcome in our model (predicting Trump or Harris as the winner based on who has higher support in each state).  Our model uses these support percentages to generate a probability of each candidate winning. This transformation helps translate fluctuating polling data into a clear prediction of the likely winner, but it also involves a level of simplification, as it assumes these support percentages accurately reflect voting intentions.

@@

It is important to note that since ‘**pct**’ stands for percentage, there may be a misconception that the values for Trump and Harris will always add up to 100%. For instance, in the summary statistics presented in Table 1, the average support for Donald Trump is 47.801%, while the average support for Kamala Harris is 47.999%. This results in a combined average support of 95.800%, indicating that there is a significant portion of respondents who are either undecided or supporting other candidates. This highlights the competitive nature of the election and emphasizes the necessity of examining not only the percentages attributed to the main candidates but also the context of voter preferences that could influence the final outcome.

### Predictive Variables
In this section, we outline the two key predictor variables used in our analysis: '**state**', which captures the geographical context of the polling data, and '**end_date**', which reflects the timing of the polls relative to the election, allowing us to assess trends in voter support over time.

#### State
The state (**state**) variable indicates the U.S. state where the poll was conducted or targeted. In the context of the U.S. presidential election, the outcome is frequently determined by several key battlegroud states. In the current 2024 election cycle, there are seven battleground states that are pivotal in the contest between Kamala Harris, representing the Democratic Party, and Donald Trump, representing the Republican Party. To enhance our model and refine our predictions, we focused exclusively on these seven battleground states. The polling results from each of these states play a crucial role in predicting electoral outcomes, as the percentage of support is evaluated on a state-by-state basis.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-state-distribution
#| fig-cap: "Distribution of Observations for Each State in Polling Data. Illustrates the number of polling observations across key battleground statesn. Notably, Pennsylvania has the highest count of observations, suggesting a more robust polling presence, while Nevada shows fewer observations. "

@@

#### End Date {#sec-end-date}
End date ('**end_date**') represents the date the poll ended, essentially the date the poll stopped collecting data from electorates. In the context of our paper, the end date of each poll serves as one of the predictive variables and is used to assess changes in voter support over time. Including end date as a variable in our model can help us identify patterns in voter support as Election Day approaches, potentially reflecting the impact of campaign events, debates, or other political developments. @fig-end-date-distribution displays the distribution of polling end dates from July 2024 to October 2024.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-end-date-distribution
#| fig-cap: "Distribution of Polling End Dates. This histogram displays the frequency of polling end dates for key battleground states, highlighting peaks in polling activity around late September."

@@

### Other Variables
This section details additional variables that provide essential context for our analysis, including **numeric grade**, which evaluates the quality of the polls; **transparency score**, indicating the openness of poll methodologies; **pollster**, identifying the organizations conducting the polls; and **start date**, which denotes the initiation of the polling period for each survey.

#### Numeric Grade
Numeric Grade ('**numeric_grade**') represents the quality of each poll, as rated by [@FiveThirtyEight], based on factors such as methodology, transparency, and historical accuracy. In the context of our paper, numeric grade allows us to weigh the reliability of different polls. By including numeric grade as a variable in our model, we aim to account for variations in poll quality, ensuring that polls with higher ratings (indicating higher reliability) contribute more significantly to our predictions of voter support. @fig-numgrade-distribution displays the distribution of numeric grades for the polls included in our analysis, ranging from 2.5 to 3.0. Polls with higher grades are generally seen as more methodologically rigorous and transparent, which helps improve the reliability of our model’s predictions. The distribution shows several peaks, indicating that certain grade levels (around 2.7, 2.8, and 3.0) are more common among the polling data we collected.

@@

```
This variability in numeric grade highlights the diversity in polling quality within our dataset. The presence of peaks suggests that some pollsters consistently receive higher ratings, likely due to their use of more rigorous methods or transparent reporting standards. Conversely, the valleys in the distribution may indicate fewer polls from sources with lower numeric grades, reflecting a potential tendency for lower-quality polls to be less prevalent or less frequently conducted. By accounting for numeric grade, we can improve the accuracy of predictions by emphasizing data from more reliable sources. This consideration is crucial, as polling quality can impact the stability of the model’s forecasts, especially in closely contested states where polling accuracy is essential for reliable predictions.

#### Transparency Score
Transparency Score ('**transparency_score**') represents the degree of methodological transparency provided by each pollster, rated on a scale from 0 to 10 by [@FiveThirtyEight]. This score reflects the extent to which pollsters disclose details about their data collection and analysis methods, such as sample size, weighting, and margin of error. In the context of our paper, transparency score is an essential variable for evaluating the reliability of polling data, as higher transparency scores generally indicate a greater level of methodological rigor and openness. @fig-transparency-score-distribution displays the distribution of transparency scores across the polling data for key battleground states. The plot shows that transparency scores cluster heavily around ratings of  7 and 9, suggesting that a majority of polls in our dataset come from sources that maintain a relatively high level of transparency. There are smaller peaks at lower scores (5 and 10), indicating that while some pollsters offer limited transparency, a few polls also achieve perfect transparency scores.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-transparency-score-distribution
#| fig-cap: "Histogram displaying the distribution of transparency scores among pollsters in the polling data. Indicates the frequency of various transparency scores, highlighting a concentration of scores around 7 and 9, which suggests a predominance of high transparency among the assessed pollsters. "

@@

#### Pollster
Pollster ('**pollster**') represents the various organizations responsible for conducting polls across key battleground states in the lead-up to the 2024 United States Presidential Election. In the context of our paper, the diversity and frequency of polling by different pollsters are essential for understanding the reliability and variance in the data. Each pollster may use different methodologies, sample populations, and data collection techniques, which can affect the results and introduce variability in polling outcomes.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-pollster-distribution
#| fig-cap: "Horizontal bar chart displaying the distribution of polls conducted by various pollsters in key battleground states. Highlights the frequency of polling activities across different organizations, with Marist and Emerson standing out as the most prolific pollsters."

@@

#### Start Date
Start date ('**start_date**') represents the date when each poll began collecting responses. In the context of our paper, start date is a key variable as it allows us to focus on polling data collected after July 21, 2024—the date Kamala Harris officially announced her presidential campaign. By filtering the dataset to include only polls with a start date after this announcement, we ensure that each poll reflects voter sentiment with Harris as an active candidate, enabling a more accurate comparison between her and Donald Trump. @fig-start-date-distribution displays the distribution of polling start dates from late July to mid-October 2024, showing when polling efforts began across key battleground states. The histogram reveals peaks in polling activity, particularly in mid-August, early September, and late September, indicating periods of heightened interest or campaign events that likely prompted increased polling.
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-start-date-distribution
#| fig-cap: "Histogram illustrating the distribution of polling start dates for key battleground states. Displays the frequency of polling activities initiated over time, with a notable peak on September 14, indicating a surge in polling efforts as the election approached."

@@

# Model {#sec-model}

## Model Overview
The Bayesian Beta regression models constructed in this paper aim to predict the probability of victory for Donald Trump and Kamala Harris in the 2024 U.S. Presidential Election by analyzing polling data of the 7 battleground states. One Bayesian Beta regression model is constructed for each candidate, and each of them estimates the proportion of support for each candidate based on polling data, accounting for both polling trends over time, and state-level differences in voter sentiment. These models allow us to evaluate how support for each candidate fluctuates over time in key states, and predict potential electoral outcomes.

@@

## Model Summary
### Trump Model Summary
@tbl-trump-model-summary displays fixed effects estimates for several key terms that capture the influence of time on his support in the battleground states. The intercept, with an estimate of -0.0923, represents the baseline log-odds of Trump’s support when other predictors are held constant. This negative intercept suggests a slightly lower baseline level of support. The term nsenddatenumdfEQ42, with an estimate of 0.0803 and a statistically significant positive confidence interval (0.0179 to 0.1426), indicates an increase in support around this time period. Similarly, nsenddatenumdfEQ44 has an estimate of 0.0789 and a confidence interval above zero (0.0014 to 0.1589), also showing a positive effect on support. Other terms such as nsenddatenumdfEQ41 and nsenddatenumdfEQ43 have less significant impacts, suggesting that certain time periods were more influential than others in shaping Trump’s support. The standard errors for the fixed effects terms range from 0.0306 to 0.0676 for nsenddatenumdfEQ43. The relatively low standard error of 0.0306 on the intercept suggests that the baseline estimate is quite stable, meaning we can be more confident about this baseline support level. However, the standard error for nsenddatenumdfEQ43 (0.0676) is higher than for other terms, indicating that this estimate is less precise.
```{r}
#| echo: false
#| eval: true
#| label: tbl-trump-model-summary
#| tbl-cap: "Coefficient estimates from the Bayesian Beta regression model for Donald Trump, showing the impact of the natural spline transformation of end date on voter support. The intercept indicates baseline log-odds, while spline terms reflect nonlinear trends in voter sentiment, with positive coefficients suggesting increased support as Election Day approaches."
#| warning: false

@@

### Harris Model Summary
@tbl-harris-model-summary similarly shows the impact of time on her support in the battleground states. The intercept for the Harris model is estimated at -0.1508, suggesting a slightly lower baseline support level compared to Trump(-0.0923). The term nsenddatenumdfEQ41 has an estimate of 0.1225 and a confidence interval 0.0553 to 0.1903, indicating a significant increase in support during this period. Another positive and significant effect is seen in nsenddatenumdfEQ43, with an estimate of 0.1191 and confidence interval from -0.0071 to 0.2417, which suggests a trend toward increased support. While other terms stay positive, they have less impacts on support. Additionally, the standard errors for fixed effects are generally smaller, ranging from 0.0279 for the intercept to 0.0631 for nsenddatenumdfEQ43. The larger standard error for nsenddatenumdfEQ43 (0.0631) suggests that there would be some uncertainty in how this time point influences support, though the positive estimate still points toward an upward trend. In comparison to the Trump model, the standard errors in the Harris model are slightly smaller for most terms, indicating potentially more consistent support trends over time in the data analyzed.
```{r}
#| echo: false
#| eval: true
#| label: tbl-harris-model-summary
#| tbl-cap: "Coefficient estimates from the Bayesian Beta regression model for Kamala Harris, illustrating the effects of the natural spline transformation of end date on voter support. The intercept represents the baseline log-odds, while the spline terms indicate nonlinear trends in voter sentiment, with positive coefficients suggesting an increase in support as Election Day nears."
#| warning: false

# Summarize fixed effects for Harris model
harris_summary <- tidy(harris_model, effects = "fixed", conf.int = TRUE) %>%
  select(term, estimate, std.error, conf.low, conf.high) %>%
  rename(
    Estimate = estimate,
    `Std. Error` = std.error,
    `Lower 95% CI` = conf.low,
    `Upper 95% CI` = conf.high
  ) %>%
  mutate(term = gsub("[^[:alnum:] ]", "", term))

@@

## State-Level Polling Averages for Trump and Harris
@tbl-outcomevar displays overview of the percentage of support for Trump and Harris by each battleground state. It provides us with a high level understanding of the current voter dynamic. The pct results are aggregated averages of all pct data points for Trump and Harris in each state. By comparing the percentage of support for T.rump and Harris, we can determine the candidate that has a leading advantage in winning each state. For example,Trump has 49.2% support in Arizona while Harris has 46.8% support, which means Trump has won the state of Arizona.
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-stateaverage
#| tbl-cap: "Percentage of support for Donald Trump and Kamala Harris across key battleground states. Trump maintains a slight lead in states such as Arizona and Georgia, while Harris shows stronger support in 5 of 7 key battleground states."

@@

## Polling Averages for Trump and Harris Over Time in Battleground States
@fig-state-polling-average provides a visual comparison of the current polling support for Donald Trump and Kamala Harris across key battleground states, tracked over time from August 2024 to October 2024. Each line plot represents the average polling percentages for both candidates in a specific state, with red lines denoting Trump’s support and blue lines representing Harris’s support. Across several states, there are noticeably increased fluctuations in polling averages for both candidates closer to the election particularly from late September to October. This pattern may indicate heightened voter interest and engagement as the election date approaches which is November 4th, 2024, it could also reflect changes in public opinion due to campaign events, political debates, or other external factors.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-polling-average
#| fig-cap: "Polling averages for Donald Trump (red) and Kamala Harris (blue) in battleground states from August to October 2024. Captures the varying fluctuations in average support percentages across Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin, indicating the competitive dynamics of the election."
#| warning: false

@@

# Plot polling averages over time for each battleground state
ggplot(polling_average, aes(x = end_date)) +
  geom_line(aes(y = Average_Trump, color = "Trump")) +
  geom_line(aes(y = Average_Harris, color = "Harris")) +
  labs(
       y = "Average Support (%)", 
       x = "Date") +
  facet_wrap(~ state) +
  scale_color_manual(values = c("Trump" = "red", "Harris" = "blue")) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +  
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1), 
    strip.text = element_text(size = 10), 
    panel.spacing = unit(0.5, "lines"),
    panel.grid.major = element_line(linewidth = 0.3) 
  )

@@

Since it is difficult to conclude a definitive predictive outcome by observing the current data, we utilize our models, described in @sec-model, to predict the possible outcomes by incorporating the given data. By using the current data, the model can identify the state-specific trends thus generating more accurate predictions of the outcome. Relevant prediction results are provided below in @fig-state-probabilities.

## State-Level Probability of Trump Winning on Election Day
By incorporating our Bayesian models, we generate the predicted probabilities of winner for both Donald Trump and Kamala Harris in key battleground states, on Election Day (**November 5th, 2024**). @fig-state-probabilities provides a visual summary of these predicted probabilities.
```{r}
#| echo: false
#| eval: true
#| label: fig-state-probabilities
#| fig-cap: "Predicted winning percentages for Donald Trump (red) and Kamala Harris (blue) in key battleground states on Election Day. Illustrates the estimated probability of victory for each candidate across Arizona, Georgia, Michigan, Nevada, North Carolina, Pennsylvania, and Wisconsin, highlighting Harris's projected advantages in the majority of these states."
#| warning: false
